The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2013-02-07 pkg/yuimadocs/inst/doc/JSS/article-new.Rnw
various update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@224 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

{\coloro koko MC here!! Why don't we make asyexp = \code{asymptotic_term} + sums? 
Also Monte-Carlo driver montecarlo = \code{simulate} with number of repetition 
+ plot + comparison of plot and theoretical curve?}

@@

\subsection{Quasi Maximum Likelihood Estimation}
{\bf[please remember: $\hat\theta$ = MLE;  $\tilde \theta$ = BAYES; $\check\theta$ = LASSO]}
\par
Consider the multidimensional diffusion process
$$
\de X_t = a(X_t,\theta_2)\de t + b(X_t,\theta_1) \de W_t
$$
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial value $X_0=x_0$. 
Quasi-MLE assumes the following approximation of the true log-likelihood for multidimensional diffusions
{\small
\begin{eqnarray}\label{qlik}
\ell_n({\bf X}_n,\theta)
=-\frac12\sum_{i=1}^n\left\{\log\text{det}(\Sigma_{i-1}(\theta_1))
+\frac{1}{\Delta_n}
\Sigma_{i-1}^{-1}(\theta_1)[\Delta X_i-\Delta_n a_{i-1}(\theta_2)]^{\otimes 2}\right\}
\end{eqnarray}}
where $\theta=(\theta_1, \theta_2)$, $\Delta X_i=X_{t_i}-X_{t_{i-1}}$, $\Sigma_i(\theta_1)=\Sigma(\theta_1,X_{t_i})$, $a_i(\theta_2)=a(X_{t_i},\theta_2)$, $\Sigma=b^{\otimes 2}$, \textcolor{red}{$A^{\otimes 2}= A A^T$} and $A^{-1}$ the inverse of $A$, $A[B]^{\otimes 2} = B^T A B$.  Then, \citep[see e.g.,][]{Yoshida92, Kessler97}, the QML estimator of $\theta$ is
\textcolor{red}{$$\hat\theta_n=\arg\max_\theta \ell_n({\bf X}_n,\theta)$$}

@@

\subsection{Adaptive Bayes Estimation}
Consider again the  diffusion process solution to
\begin{equation} 
\de X_t=a(X_t,\theta_2)\de t+b(X_t,\theta_1)\de W_t,
\end{equation}
and the quasi likelihood defined in \eqref{qlik}.

@@

\subsection{Change-Point Analysis}
Consider a multidimensional stochastic differential equation of the form
$$
\de Y_t = a_t \de t + b(X_t,\theta) \de W_t,\ \ t\in[0,T], 
$$
where $W_t$ \textcolor{red}{is an} $r$-dimensional Wiener process and $a_t$ and $X_t$ are multidimensional processes and $b$ is the diffusion coefficient (volatility) matrix.
When $Y=X$ the problem is a diffusion model.
The process $a_t$ may have jumps but should not explode and it is treated as a nuisance in this model.
The change-point problem for the volatility is formalized as follows
$$
Y_t=
\Bigg\{
\begin{array}{ll}
Y_0+\int_0^t a_s \de s+\int_0^t b(X_s,\theta_0^*) \de W_s
& \mbox{ for } t\in[0,\tau^*)
\\
Y_{\tau^*}+\int_{\tau^*}^t a_s \de s+\int_{\tau^*}^t b(X_s,\theta_1^*) \de W_s
& \mbox{ for } t\in[\tau^*,T]. 
\end{array}
$$
The change-point $\tau^*$ instant is unknown and 
is  to be estimated, along with $\theta_0^*$ and $\theta_1^*$, from the observations sampled 
from the path of $(X,Y)$. The \pkg{yuima} implements the quasi-maximum likelihood approach as in \citet{iacyos09} described in the following.
Let  $\Delta_iY=Y_{t_i}-Y_{t_{i-1}}$ and 
define 
\begin{equation} 
\Phi_n(t;\theta_0,\theta_1)
=
\sum_{i=1}^{[nt{/T}]}G_i(\theta_0)+\sum_{i={[nt{/T}]+1}}^nG_i(\theta_1),
\label{eq:phiCH10}
\end{equation}
with 
\begin{equation} 
G_i(\theta) =
\log\det S(X_{t_{i-1}},\theta)
+ \Delta_n^{-1}(\Delta_iY)'S(X_{t_{i-1}},\theta)^{-1}(\Delta_iY)
\label{eq:GiCH10}
\end{equation}
\textcolor{red}{and $S=\sigma^{\otimes 2}$.}
Suppose that there exists an estimator 
$\hat{\theta}_k$ for each $\theta_k$, $k=0,1$. 
In case $\theta_k^*$ are known, we define $\hat{\theta}_k$ 
just as $\hat{\theta}_k=\theta_k^*$. 
The change-point estimator  of $\tau^*$ is
\begin{eqnarray*} 
\hat{\tau}_n&=&\arg\!\!\min\limits_{t\in[0,T]} 
\Phi_n(t;\hat{\theta}_0,\hat{\theta}_1).
\end{eqnarray*}

\subsubsection{Example  of Volatility Change-Point Estimation for 2-Dimensional SDE's}
One example of model that can be analyzed by this technique is, for example, the 2-dimensional stochastic differential equation
$$
\left(\begin{array}{c}
\de X_{1,t}\\ \de X_{2,t} 
\end{array}\right)
= 
\left(\begin{array}{c}
b_1(X_{1,t}) \\
b_2(X_{2,t}) \\ 
\end{array}\right) \de t
+
\left(\begin{array}{cc}
\theta_{1.k}  \cdot  X_{1,t}&0  \cdot  X_{1,t} \\ 
0  \cdot  X_{2,t}&\theta_{2.k}  \cdot  X_{2,t} \\ 
\end{array}\right)
\left(\begin{array}{c}
\de W_{1,t}\\ \de W_{2,t} 
\end{array}\right),\quad \textcolor{red}{ t\in[0,T],}
$$
where $b_1(\cdot)$ and $b_2(\cdot)$ are any functions and $\theta_{1.k}$ and $\theta_{2.k}$ the value of the parameters before ($k=0$) and after $k=1$) the change-point. 
Just for simplicity and in order to simulate some data, we specify some specific form for $b_1(\cdot)$ and $b_2(\cdot)$ but this information will not be used in the change-point analysis.
For example, we will simulate
the following 2-dimensional stochastic differential equation
$$
\left(\begin{array}{c}
 \de X_{1,t}\\ \de X_{2,t} 
 \end{array}\right)
 = 
\left(\begin{array}{c}
 \sin(X_{1,t}) \\
 3 - X_{2,t} \\ 
 \end{array}\right) \de t
 +
\left(\begin{array}{cc}
 \theta_{1.k}  \cdot  X_{1,t}&0  \cdot  X_{1,t} \\ 
 0  \cdot  X_{2,t}&\theta_{2.k}  \cdot  X_{2,t} \\ 
 \end{array}\right)
\left(\begin{array}{c}
 \de W_{1,t}\\ \de W_{2,t} 
 \end{array}\right),\quad \textcolor{red}{ t\in[0,T],}
$$
$$
X_{1,0}=1.0,\quad
X_{2,0}=1.0,\quad
$$
with change-point instant at time $\tau=0.4$ \textcolor{red}{and $T=10$}.   First, we describe the model to be simulated
<<cpoint1, echo=TRUE,results=hide>>=
diff.matrix <- matrix(c("theta1.k*x1","0*x2","0*x1","theta2.k*x2"), 2, 2)
drift.c <- c("sin(x1)", "3-x2")
drift.matrix <- matrix(drift.c, 2, 1)
ymodel <- setModel(drift=drift.matrix, diffusion=diff.matrix, time.variable="t",
state.variable=c("x1", "x2"), solve.variable=c("x1", "x2"))
@
and then simulate two trajectories. One up to the change-point $\tau=4$ with parameters \textcolor{red}{$\theta_{1.0}=0.5$ and  $\theta_{2.0}=0.3$}
<<cpoint3,results=hide>>=
n <- 1000

@@

\subsection{LASSO Model Selection}
Let $X_t$ be a  diffusion process solution to
$$
\de X_t = a( X_t,\alpha) \de t + b(X_t,\beta)  \de W_t
$$
$$\alpha=(\alpha_1,...,\alpha_{p})'\in\Theta_p\subset \mathbb{R}^p, \quad p\geq1$$
$$\beta=(\beta_1,...,\beta_q)'\in\Theta_q\subset \mathbb{R}^q, \quad q\geq1$$
with $b:\Theta_p\times\mathbb{R}^d\to \mathbb{R}^d$, $\sigma:\Theta_q\times \mathbb{R}^d\to \mathbb{R}^d\times \mathbb{R}^m$ and $W_t,t\in [0,T]$, is a standard Brownian motion in $\mathbb{R}^m$. 
We assume that the functions $a$ and $b$ are known up to   $\alpha$ and $\beta$.
We denote by $\theta=(\alpha,\beta)\in\Theta_p\times \Theta_q=\Theta$ the parametric vector and with $\theta_0=(\alpha_0,\beta_0)$ its unknown true value. 
Let $\mathbb{H}_n({\bf X}_n,\theta) = \ell_n({\bf X}_n,\theta)$ from equation \eqref{qlik}.
The quasi-MLE $\tilde{\theta}_n$ for this model is the solution of the following problem
$$
\tilde{\theta}_n=(\tilde\alpha_n,\tilde\beta_n)'=\arg\min_\theta \mathbb{H}_n({\bf X}_n,\theta)
$$ 
The adaptive LASSO estimator is defined as the solution to the quadratic problem under $L_1$ constraints
$$
\hat{\theta}_n=(\hat\alpha_n,\hat\beta_n)=\arg\min_\theta\mathcal{F}(\theta).
$$
with
$$
\mathcal{F}(\theta)=(\theta-\tilde{\theta}_n)^T\ddot{\mathbb{H}}_n({\bf X}_n, \tilde\theta_n)(\theta-\tilde{\theta}_n)+\sum_{j=1}^p\lambda_{n,j}|\alpha_j| +\sum_{k=1}^q\gamma_{n,k}|\beta_k|
$$
For more details see \citet{DegIac10b}.
The tuning parameters should be chosen as in \citet{Zou06} in the following way
\begin{equation}
\label{eq:penalty}
\lambda_{n,j} = \lambda_0 |\tilde \alpha_{n,j}|^{-\delta_1}, \qquad
\gamma_{n,k} = \gamma_0 |\tilde \beta_{n,j}|^{-\delta_2}
\end{equation}
where $\tilde \alpha_{n,j}$ and  $\tilde \beta_{n,k}$ are the unpenalized QML estimator of $\alpha_j$ and $\beta_k$ respectively, $\delta_1, \delta_2>0$ and usually taken unitary.

>>>>

@@

\subsection{Quasi Maximum Likelihood Estimation}
{\bf ADD SOMETHING ON ASYMPT DISTRIBUTION \& SAMPLING SCHEME} 
Consider the multidimensional diffusion process
\begin{equation}
\label{eq:sdemle}
\de X_t = a(X_t,\theta_2)\de t + b(X_t,\theta_1) \de W_t
\end{equation}
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial value $X_0=x_0$. 
Moreover, $\theta_1\in\Theta_1\subset \mathbb{R}^p, p\geq1$,
$\theta_2\in\Theta_2\subset \mathbb{R}^q, q\geq1$,
$a:\mathbb{R}^d\times\Theta_2 \to \mathbb{R}^d$ and $b:\mathbb{R}^d\times\Theta_1 \to \mathbb{R}^d\times \mathbb{R}^r$.

Quasi-MLE assumes the following approximation of the true log-likelihood for multidimensional diffusions
{\small
\begin{eqnarray}\label{qlik}
\ell_n({\bf X}_n,\theta)
=-\frac12\sum_{i=1}^n\left\{\log\text{det}(\Sigma_{i-1}(\theta_1))
+\frac{1}{\Delta_n}
\Sigma_{i-1}^{-1}(\theta_1)[\Delta X_i-\Delta_n a_{i-1}(\theta_2)]^{\otimes 2}\right\}
\end{eqnarray}}
where $\theta=(\theta_1, \theta_2)$, $\Delta X_i=X_{t_i}-X_{t_{i-1}}$, $\Sigma_i(\theta_1)=\Sigma(\theta_1,X_{t_i})$, $a_i(\theta_2)=a(X_{t_i},\theta_2)$, $\Sigma=b^{\otimes 2}$, \textcolor{red}{$A^{\otimes 2}= A A^T$} and $A^{-1}$ the inverse of $A$, $A[B]^{\otimes 2} = B^T A B$.  Then, \citep[see e.g.,][]{Yoshida92, Kessler97}, the QML estimator of $\theta$ is
\textcolor{red}{$$\hat\theta=\arg\max_\theta \ell_n({\bf X}_n,\theta)$$}

@@

\subsection{Adaptive Bayes Estimation}
Consider again the  diffusion process solution to \eqref{eq:sdemle}
and the quasi likelihood defined in \eqref{qlik}.

@@

\subsection{Change-Point Analysis}
Consider a multidimensional stochastic differential equation of the form
$$
\de Y_t = a_t \de t + b(X_t,\theta) \de W_t,\ \ t\in[0,T], 
$$
where $W_t$ \textcolor{red}{is an} $r$-dimensional Wiener process and $a_t$ and $X_t$ are multidimensional processes, $\theta\in\Theta\subset \mathbb{R}^p$,  $b:\mathbb{R}^d\times\Theta \to \mathbb{R}^d\times \mathbb{R}^r$, is the diffusion coefficient (volatility) matrix.

When $Y=X$ the problem is a diffusion model.
The process $a_t$ may have jumps but should not explode and it is treated as a nuisance in this model.
The change-point problem for the volatility is formalized as follows
$$
Y_t=
\Bigg\{
\begin{array}{ll}
Y_0+\int_0^t a_s \de s+\int_0^t b(X_s,\theta_0^*) \de W_s
& \mbox{ for } t\in[0,\tau^*)
\\
Y_{\tau^*}+\int_{\tau^*}^t a_s \de s+\int_{\tau^*}^t b(X_s,\theta_1^*) \de W_s
& \mbox{ for } t\in[\tau^*,T]. 
\end{array}
$$
The change-point $\tau^*$ instant is unknown and 
is  to be estimated, along with $\theta_0^*$ and $\theta_1^*$, from the observations sampled 
from the path of $(X,Y)$. The \pkg{yuima} implements the quasi-maximum likelihood approach as in \citet{iacyos09} described in the following.
Let  $\Delta_iY=Y_{t_i}-Y_{t_{i-1}}$ and 
define 
\begin{equation} 
\Phi_n(t;\theta_0,\theta_1)
=
\sum_{i=1}^{[nt{/T}]}G_i(\theta_0)+\sum_{i={[nt{/T}]+1}}^nG_i(\theta_1),
\label{eq:phiCH10}
\end{equation}
with 
\begin{equation} 
G_i(\theta) =
\log\det S(X_{t_{i-1}},\theta)
+ \Delta_n^{-1}(\Delta_iY)'S(X_{t_{i-1}},\theta)^{-1}(\Delta_iY)
\label{eq:GiCH10}
\end{equation}
\textcolor{red}{and $S=b^{\otimes 2}$.}
Suppose that there exists an estimator 
$\hat{\theta}_k$ for each $\theta_k$, $k=0,1$. 
In case $\theta_k^*$ are known, we define $\hat{\theta}_k$ 
just as $\hat{\theta}_k=\theta_k^*$. 
The change-point estimator  of $\tau^*$ is
\begin{eqnarray*} 
\hat{\tau}&=&\arg\!\!\min\limits_{t\in[0,T]} 
\Phi_n(t;\hat{\theta}_0,\hat{\theta}_1).
\end{eqnarray*}

\subsubsection{Example  of Volatility Change-Point Estimation for 2-Dimensional SDE's}
One example of model that can be analyzed by this technique is, for example, the 2-dimensional stochastic differential equation
$$
\left(\begin{array}{c}
\de X_{1,t}\\ \de X_{2,t} 
\end{array}\right)
= 
\left(\begin{array}{c}
a_1(X_{1,t}) \\
a_2(X_{2,t}) \\
\end{array}\right) \de t
+
\left(\begin{array}{cc}
\theta_{1.k}  \cdot  X_{1,t}&0  \cdot  X_{1,t} \\ 
0  \cdot  X_{2,t}&\theta_{2.k}  \cdot  X_{2,t} \\ 
\end{array}\right)
\left(\begin{array}{c}
\de W_{1,t}\\ \de W_{2,t} 
\end{array}\right),\quad \textcolor{red}{ t\in[0,T],}
$$
where $b_1(\cdot)$ and $b_2(\cdot)$ are any functions and $\theta_{1.k}$ and $\theta_{2.k}$ the value of the parameters before ($k=0$) and after $k=1$) the change-point. 
Just for simplicity and in order to simulate some data, we specify some specific form for $b_1(\cdot)$ and $b_2(\cdot)$ but this information will not be used in the change-point analysis.
For example, we will simulate
the following 2-dimensional stochastic differential equation
$$
\left(\begin{array}{c}
 \de X_{1,t}\\ \de X_{2,t} 
 \end{array}\right)
 = 
\left(\begin{array}{c}
 \sin(X_{1,t}) \\
 3 - X_{2,t} \\ 
 \end{array}\right) \de t
 +
\left(\begin{array}{cc}
 \theta_{1.k}  \cdot  X_{1,t}&0  \cdot  X_{1,t} \\ 
 0  \cdot  X_{2,t}&\theta_{2.k}  \cdot  X_{2,t} \\ 
 \end{array}\right)
\left(\begin{array}{c}
 \de W_{1,t}\\ \de W_{2,t} 
 \end{array}\right),\quad \textcolor{red}{ t\in[0,T],}
$$
$$
X_{1,0}=1.0,\quad
X_{2,0}=1.0,\quad
$$
with change-point instant at time $\tau=0.4$ \textcolor{red}{and $T=10$}.   First, we describe the model to be simulated
<<cpoint1, echo=TRUE,results=hide>>=
diff.matrix <- matrix(c("theta1.k*x1","0*x2","0*x1","theta2.k*x2"), 2, 2)
drift.c <- c("sin(x1)", "3-x2")
drift.matrix <- matrix(drift.c, 2, 1)
ymodel <- setModel(drift=drift.matrix, diffusion=diff.matrix, time.variable="t",
state.variable=c("x1", "x2"), solve.variable=c("x1", "x2"))
@
and then simulate two trajectories. One up to the change-point $\tau=4$ with parameters \textcolor{red}{$\theta_{1.0}=0.5$ and  $\theta_{2.0}=0.3$}
<<cpoint3,results=hide>>=
n <- 1000

@@

\subsection{LASSO Model Selection}
Let $X_t$ be a  diffusion process solution to
$$
\de X_t = a( X_t,\alpha) \de t + b(X_t,\beta)  \de W_t
$$
$$\alpha=(\alpha_1,...,\alpha_{p})'\in\Theta_p\subset \mathbb{R}^p, \quad p\geq1$$
$$\beta=(\beta_1,...,\beta_q)'\in\Theta_q\subset \mathbb{R}^q, \quad q\geq1$$
with $b:\Theta_p\times\mathbb{R}^d\to \mathbb{R}^d$, $\sigma:\Theta_q\times \mathbb{R}^d\to \mathbb{R}^d\times \mathbb{R}^m$ and $W_t,t\in [0,T]$, is a standard Brownian motion in $\mathbb{R}^m$. 
We assume that the functions $a$ and $b$ are known up to   $\alpha$ and $\beta$.
We denote by $\theta=(\alpha,\beta)\in\Theta_p\times \Theta_q=\Theta$ the parametric vector and with $\theta_0=(\alpha_0,\beta_0)$ its unknown true value. 
Let $\mathbb{H}_n({\bf X}_n,\theta) = \ell_n({\bf X}_n,\theta)$ from equation \eqref{qlik}.
The quasi-MLE $\hat{\theta}$ for this model is the solution of the following problem
$$
\hat{\theta}=(\hat\alpha,\hat\beta)'=\arg\min_\theta \mathbb{H}_n({\bf X}_n,\theta)
$$ 
The adaptive LASSO estimator is defined as the solution to the quadratic problem under $L_1$ constraints
$$
\check{\theta}=(\check\alpha,\check\beta)=\arg\min_\theta\mathcal{F}(\theta).
$$
with
$$
\mathcal{F}(\theta)=(\theta-\hat{\theta})^T\ddot{\mathbb{H}}_n({\bf X}_n, \hat\theta)(\theta-\hat{\theta})+\sum_{j=1}^p\lambda_{n,j}|\alpha_j| +\sum_{k=1}^q\gamma_{n,k}|\beta_k|
$$
For more details see \citet{DegIac10b}.
The tuning parameters should be chosen as in \citet{Zou06} in the following way
\begin{equation}
\label{eq:penalty}
\lambda_{n,j} = \lambda_0 |\hat \alpha_{j}|^{-\delta_1}, \qquad
\gamma_{n,k} = \gamma_0 |\hat \beta_{j}|^{-\delta_2}
\end{equation}
where $\hat \alpha_{j}$ and  $\hat \beta_{k}$ are the unpenalized QML estimator of $\alpha_j$ and $\beta_k$ respectively, $\delta_1, \delta_2>0$ and usually taken unitary.
